Left-linear rules in adhesive categories

### 2024-02-13 icalp/fusione.tex
minor

So far, most of the theoretical results focussed on what are called \emph{linear rules}: 
intuitively, a step is required only to consume, preserve and generate entities. 
However, in some situations it is also necessary to ``merge'' parts of the state. 
This happens already in the context of string diagram mentioned above, but also 
in graphical implementations of nominal calculi where, as a result of name passing, 
the received name is identified with a local one at the
receiver~\cite{CVY:ESSPE,Gad07} or in the visual modelling of bonding in
biological/chemical processes~\cite{PUY:MBPE}. A similar mechanisms is
at the core of e-graphs (equality graphs), recently used for
rewrite-driven compiler optimisations~\cite{WNW:egg}: rather than
explicitly merging nodes corresponding to equal expression, the idea
here -- conceptually and, as we will see, also mathematically similar --
is to maintain an equivalence %\todo{P: more precisely a congruence}
over the nodes.  
%\todo{P: la sequenza di lavori su questo \`e ampia,
%vedere se c'\`e qualcosa d'altro}

Technically, to acquire the expressiveness needed for capturing
fusions, i.e. the merge of some elements in the state, requires to
move from linear to left-linear rewriting rules. While left-linear
rules have been considered in various instances (mainly in categories
of graphs) and some results concerning the corresponding rewriting
theory have been put forward, to the best of our knowledge the
phenomena arising when devising a theory of indepedence and
concurrency for rewriting with left-linear rules in the general
setting of adhesive and quasi-adhesive categories
%(which capture most ``applications'') 
have not been systematically
explored~\cite{Ehrig1976,EHP:BRfTToHLRS,baldan2011adhesivity} \todo{P: see also \cite{HMP:DPORADD,BHK:CTNL} ... verificare e aggiungere}

@@

There is, however, a further problem.  For linear
rewriting systems, switch equivalence can be characterised in terms of the
existence of what is called a consistent permutation~\cite{xxx}.
%(roughly, an isomorphism between their colimits). 
This result, which
is folklore for various categories of graphs, is explicitly proved
here for $\mathcal{M}$-adhesive categories.
%
The characterisation of switch equivalence based on consistent
permutations is the key to derive some
properties for linear rewriting systems, namely
\begin{enumerate}
\item \emph{consistency of switching}: when transforming
  sequences of rewriting steps into equivalent ones by switching independent steps, the
  result not depends on the order in which switchings are applied, but
  only on the resulting permutation;
\item \emph{globality of independence}: if two independent steps are
  moved forward or backward due to the application of switchings to a
  derivation, they remain independent.
\end{enumerate}
Properties (1) and (2) allow one to deduce a canonical form for equivalence proofs 
between sequences of rewriting steps seen as concurrent computations.

Unfortunately, for left-linear rewriting systems the characterisation
of switch equivalence based on consistent permutations fails: the
existence of a consistent permutation is only necessary but not
sufficient, hence it cannot be used for proving (1) and (2).
%
This leads to the question asking whether the properties above hold.
We show that actually much of the theory can be retained: consistency
of switching still holds, while globality of independence holds in a
weak form, conceptually linked to the fact that fusions introduce a
form of disjunctive causality. Finally, we prove that a canonical form
for switching sequences can also be recovered.

@@

  As a consequence, $\der{D}$ and $\der{D}'$ are switch equivalent, but
  this is not acceptable when viewing equivalence classes of
  derivations as concurrent computations: in $\der{D}$ the first two
  steps are not sequential independent, while in $\der{D}'$ they are,
  intuitively because in $\der{D}$ rule $\rho_1$ uses the node
  generated by $\rho_0$, while in $\der{D}'$ rule $\rho_1$ uses the
  node which was in the initial graph. From the technical point of
  view, the property of being switch equivalent is not closed by
  prefix, something which prevents to derive a sensible concurrent
  sematics: if we limit to the first two steps, derivations
  $\der{D}_0; \der{D}_1$ and $\der{D}_0'; \der{D}_1'$ are not switch
  equivalent while $\der{D}$ and $\der{D}'$ would be
\end{example}

Moreover, limiting sequential independence to the case in which the independence
pair is unique (as suggested in~ ...) brings to the same problems.

@@

\begin{definition}[well switching rewriting systems]
A left-linear DPO rewriting system $(\X, \R)$ is \emph{well switching} if it is strong enforcing and, for every derivation $\der{D}:=\{\dder{D}_{i}\}_{i=0}^1$, there is at most one independence paire between $\dder{D}_0$ and $\dder{D}_1$.
\end{definition}

The previous definition might sound restrictive. However, it is immediate to see that linear rewriting systems are all well switching (see Proposition~\ref{pr:weak}).
%
Moreover, we next introduce a class of rewriting systems, comprising
all the ones used in concrete modelling applications of DPO rewriting
(e.g., in the encoding of process calculi ..., ...) are \todo{INSERIRE ESEMPI QUI}
actually well switching.

@@

The first lemma capture the essence of the confluence of
switchings. It shows that if we have a derivation consisting of three
rewriting steps, say $r_0\, r_1\, r_2$ and we switch them leading to the
sequence $r_2\, r_1\, r_0$, we obtain the same resulting derivation,
independently from the order of the switchings.

@@

There is another
useful properties of weakly sequentially independent direct
derivations which is worth mentioning.

>>>>

So far, most of the theoretical results focussed on what are called
\emph{linear rules}: intuitively, a step is required only to consume,
preserve and generate entities.  However, in some situations it is
also necessary to ``merge'' parts of the state.  This happens already
in the context of string diagram mentioned above, but also in
graphical implementations of nominal calculi where, as a result of
name passing, the received name is identified with a local one at the
receiver~\cite{CVY:ESSPE,Gad07} or in the visual modelling of bonding
in biological/chemical processes~\cite{PUY:MBPE}. A similar mechanisms
is at the core of e-graphs (equality graphs), recently used for
rewrite-driven compiler optimisations~\cite{WNW:egg}: rather than
explicitly merging nodes corresponding to equal expression, the idea
here -- conceptually and, as we will see, also mathematically similar
-- is to maintain an
equivalence %\todo{P: more precisely a congruence}
over the nodes.

Technically, to acquire the expressiveness needed for capturing
fusions, i.e. the merge of some elements in the state, requires to
move from linear to left-linear rewriting rules. While left-linear
rules have been considered in various instances (mainly in categories
of graphs) and some results concerning the corresponding rewriting
theory have been put forward, to the best of our knowledge the
phenomena arising when devising a theory of indepedence and
concurrency for rewriting with left-linear rules in the general
setting of adhesive and quasi-adhesive categories
%(which capture most ``applications'') 
have not been systematically
explored~\cite{Ehrig1976,EHP:BRfTToHLRS,baldan2011adhesivity},
even if an investigation in the alternative setting of quasi-topoi has 
been presented in~\cite{BehrHK23}.

@@

% \begin{enumerate}
% \item \emph{confluence of switching}: when transforming derivation
%   sequences into equivalent ones by switching independent steps, the
%   result not depends on the order in which switchings are applied, but
%   only on the resulting permutation;

% \item \emph{globality of independence}: if two independent steps are
%   moved forward or backward due to the application of switchings to a
%   derivation, they remain independent.
% \end{enumerate}

There is, however, a further problem. In the literature on linear
rewriting systems, a central result is the a characterisation of
switch equivalence of sequences of steps based on what are called processes or
consistent permutations~\cite{Handbook, tobias}.\todo{Handbook, graph processes, tobias}
In turn this is the key
to derive some simple but fundamental properties for linear rewriting
systems, namely

\begin{enumerate}
\item \emph{consistency of switching}: when transforming
  sequences of rewriting steps into equivalent ones by switching independent steps, the
  result not depends on the order in which switchings are applied, but
  only on the resulting permutation;
\item \emph{globality of independence}: if two independent steps are
  moved forward or backward due to the application of switchings to a
  sequence, they remain independent.
\end{enumerate}
Properties (1) and (2) allow one to deduce a canonical form for equivalence proofs 
between sequences of rewriting steps seen as concurrent computations.

Unfortunately, for left-linear rewriting systems the characterisation
of switch equivalence based on consistent permutations fails: the
existence of a consistent permutation is only necessary but not
sufficient, hence it cannot be used for proving (1) and (2).
%
This leads to asking whether the properties above hold.
We show that actually much of the theory can be retained: consistency
of switching still holds, while globality of independence holds in a
weak form, conceptually linked to the fact that fusions introduce a
form of disjunctive causality. Finally, we prove that a canonical form
for switching sequences can also be recovered.

@@

  As a consequence, $\der{D}$ and $\der{D}'$ are switch equivalent, but
  this is not acceptable when viewing equivalence classes of
  derivations as concurrent computations: in $\der{D}$ the first two
  steps are not sequential independent, while in $\der{D}'$ they are,
  intuitively because in $\der{D}$ rule $\rho_1$ uses the node
  generated by $\rho_0$, while in $\der{D}'$ rule $\rho_1$ uses the
  node which was in the initial graph. From the technical point of
  view, the property of being switch equivalent is not closed by
  prefix, something which prevents to derive a sensible concurrent
  sematics: if we limit to the first two steps, derivations
  $\dder{D}_0; \dder{D}_1$ and $\dder{D}_0'; \dder{D}_1'$ are not switch
  equivalent while $\der{D}$ and $\der{D}'$ would be
\end{example}

\full{Moreover, limiting sequential independence to the case in which the independence
pair is unique (as suggested in~\cite{baldan2017domains}) brings to the same problems.}

@@

\begin{definition}[Well switching rewriting systems]
A left-linear DPO rewriting system $(\X, \R)$ is \emph{well switching} if it is strong enforcing and, for every derivation $\der{D}:=\{\dder{D}_{i}\}_{i=0}^1$, there is at most one independence pair between $\dder{D}_0$ and $\dder{D}_1$.
\end{definition}

The previous definition might sound restrictive. However, it is immediate to see that linear rewriting systems are all well switching (see Proposition~\ref{pr:weak}).
%
Moreover, we next introduce a class of rewriting systems, comprising
all the ones used in concrete modelling applications of DPO rewriting
(e.g., in the encoding of process calculi ..., ...) are \todo{INSERIRE
  ESEMPI QUI} actually well switching.
%
Roughly speakings, these are possibly hierarchical graphical structures where rules are contrained not to merge elements of top level sorts in the hiearchy (for graphs, nodes can be merged while edges can not).

@@

The second lemma capture the essence of the confluence of
switchings. It shows that if we have a derivation consisting of three
rewriting steps, say $\dder{D}_0\cdot \dder{D}_1\cdot  \dder{D}_2$ and we switch them leading to the
sequence $\dder{D}_2\cdot  \dder{D}_1\cdot  \dder{D}_0$, we obtain the same resulting derivation,
independently from the order of the switchings.

@@

There is another
useful properties of  sequentially independent direct
derivations which is worth mentioning.\todo{A: va tolto weakly?}
